Combinatorial interpretations of holonomic recurrences

### 2021-08-22 article/main.tex
Minor fixes

\(u, v\) are implicitly assumed to be a nodes.

@@

We say that node \(v\) is an oneling \(\textiff\) \(v\) has no sibling. Note that the root is oneling even if it has no parent.

@@

We say that node \(v\) is an virtual-oneling \(\textiff\) \(v\) has no sibling being virtual node.

@@

\(\n\) is implicitly defined as a default identifier a node.

\(\no\) is implicitly defined as a identifier of non-existing node (eg.\ if \(v\) is a leaf, \(\tleft(v) = \n\)).

@@

We say that node \(v\) is right (left) \(\textiff\) \(v\) is a right (left) child of its parent.

We say that node \(v\) has a right (left) parent \(u\) \(\textiff\) \(v\) is a left (right) child of \(u\).

@@

\(x_i\) is implicitly assumed to be \(i\)-th element of the sequence \(X\), where \(x\) can be any lower-case letter and \(X\) is upper-case \(x\).

@@

We will be analyzing structures that can be expressed using recursive expressions. Eg.\ we can define full binary tree recursively as a node having no children (ie.\ being a leaf) or having exactly two subtrees being full binary trees. To express the alternative in the construction rule, we use \(+\). The reason will become obvious soon.

@@

There exist multiple well-known methods and tools that allow us to work with generating functions without need of digging into details. The tool we will be using extensively in this work is Maple package called GFUN\cite{gfun}.

@@

\begin{definition}
    \label{def:holo_2}
    \cite{complexity}
    Equivalently to definition\ \ref{def:holo_1}, \(A = (a_i)_{i=0}^{\infty}\) is holonomic \(\textiff\) exists generating function in the following form:
    \[\sum_{i=0}^{r} p_i(z) \gf{A}^{(i)}(z) = b(z)\]
    for some fixed \(r \in \N\); \(b, p_0, \ldots, p_r \in \poly\) not all identically zero.
\end{definition}

@@

In the article we try to gather methods and results that may be helpful for anyone who will try to find such interpretation, as we believe that such interpretation may give better insight into the nature of the structure and make it easier to study its behaviour.

@@

We define the size of tree as a number of internal nodes (ie.\ nodes that are not leaves). Our generating function is therefore:

@@

This recurrence describes also Catalan numbers.

It is easy to see that tree of size \(n\) (\(n\) internal nodes) has \(n + 1\) leaves and therefore \(2n + 1\) nodes in total. Analogically the tree of size \(n + 1\) has \(n + 2\) leaves and \(2n + 3\) nodes in total. If you see some resemblance to our holonomic recurrence, you are perfectly right.

@@

First case can be also split into 2 subcases. Without loss of the generality we can analyze just one of the operations, let it be \(R_l\). The second case will be symmetric.
\begin{itemize}
    \item for some nodes \(u \neq v\), \(R_l(T, u) = R_l(T, v)\). Let \(i_u, i_v\) be the indices of respectively \(\enc{u}\) and \(\enc{v}\) in \(\enc{T}\). Let \(\pointed{i}_u, \pointed{i}_v\) be the indices of pointed node respectively in \(\enc{R_l(T, u)}\) and \(\enc{R_l(T, v)}\). Due to the definition of \(R_l\), we get that \(\pointed{i}_u = i_u + 1\) and \(\pointed{i}_v = i_v + 1\), which implies that \(u = v\) what we assumed false.
    \item for some trees \(T \neq S\), \(R_l(T, u) = R_l(S, v)\). Let \(i_u, i_v\) be the indices of respectively \(\enc{u}\) in \(\enc{T}\) and \(\enc{v}\) in \(\enc{S}\). Let \(\pointed{i}_u, \pointed{i}_v\) be the indices of pointed node respectively in \(\enc{R_l(T, u)}\) and \(\enc{R_l(S, v)}\). If \(\pointed{i}_u \neq \pointed{i}_v\), this case is trivially false, so we can assume that \(\pointed{i}_u = \pointed{i}_v\). Therefore, again from the definition of \(R_l\), we have that \(i_u = i_v\). That implies that \(\enc{\tPre(T, u)} \enc{\tSub(u)} \enc{\tSuf(T, u)} = \enc{\tPre(S, v)} \enc{\tSub(v)} \enc{\tSuf(S, v)} \), which then sums up to \(\enc{T} = \enc{S}\), which we assumed false.
\end{itemize}

@@

The next case we will show is case of Schröder trees\cite{schroder}.

@@

We can also move \(\gf{T}(z)\) to the other side, ie.:

@@

Let's define 3 transformations: \(L_1\), \(L_2\) and \(R_1\).

\begin{figure}[h]
    \begin{center}
    \includeinlinesvg{.16}{schroder__lr_base}%
    \(\xRightarrow{L_1}\)%
    \includeinlinesvg{.16}{schroder__l1}%
    \hspace{.1\textwidth}%
    \includeinlinesvg{.16}{schroder__lr_base}%
    \(\xRightarrow{L_2}\)%
    \includeinlinesvg{.16}{schroder__l2}%

    \includeinlinesvg{.16}{schroder__lr_base}%
    \(\xRightarrow{R_1}\)%
    \includeinlinesvg{.16}{schroder__r1}%
    \end{center}
    \caption{Transformations \(L_1\), \(L_2\) and \(R_1\)}
    \label{fig:transformations_L1_L2_R1}
\end{figure}

@@

As you could see in the examples above, most of the solutions base on the simple analysis of number of laves, internal nodes and other similar properties. In this section, we've gathered some methods that may be helpful when trying to analyze the problem of interpreting combinatorially some holonomic recurrence.

@@

First thing that we have to select is defining the size carefully. We have few standard choices of computing tree sizes:
\begin{itemize}
    \item counting all existing nodes,
    \item counting internal nodes,
    \item counting leaves,
    \item counting null-nodes (ie.\ children of the leaves),
    \item edges.
\end{itemize}

It may seem counter-intuitive at first, that this changes anything crucial, but the holonomic recurrence may change drastically making the problem more difficult than it should be.

There are 2 most common symptoms that we may want to choose different method of counting the size:
\begin{itemize}
    \item we get empty classes of the objects, eg.\ if we analyze problem of full binary trees and define size as number of all existing nodes, we can not generate tree of even size, therefore half of our classes are empty;
    \item we want to join trees using common root and we get many \(\pm1\) in our notes. Note that if you count internal nodes, such catenation creates tree of size being sum of sizes of the components increases by 1. If you had defined the size as number of leaves or null-nodes, the size would be just plain sum of sizes of the smaller trees.
\end{itemize}

@@

Another method that may help finding the interpretation is inductive translation\cite{doron}. The method is based on the idea that most of the time there is easy to find some natural induction (not necessarily holonomic) that has trivial interpretation in bijective language.

@@

Using our inductive hypothesis, ie.\ holonomic recurrence and moving problematic cases out from the sums (in that case \(B(1)\) needs to be analyzed separately, as we did not define binary trees with no leaves), we can get the right side to use \(B(n - 1)\) and prove that the holonomic recurrence using non-linear one. The only part that is left is to translate this proof to the bijective language, which in such simple case as binay trees is almost trivial, especially if one has seen the solution once.

@@

If all elaborate methods has failed and the problem is still too big to interpret it by hand, a computer comes handy.

@@

The definition above is quite popular, but introduces several problems. One of them is problem of potential collision of lambda-abstracted variables and difficulty of finding to which lambda abstraction is the variable is bound (or is it even free variable, ie.\ not bound). To address that, we can use de Bruijn indices.

@@

\begin{definition}
    \(s\) is a node with just right child.
\end{definition}

\begin{definition}
    \(l\) is a node with just left child.
\end{definition}

@@

We also created a framework\cite{transformator}, called transformator, that may help visualize our work. It is based on the idea that every transformation described using bijective language can be translated to a function transforming node and finite number of its surrounding nodes. This function can not discriminate nodes on the grounds of their kind. The only discrimination that may occur may be connected to the following cases:
\begin{itemize}
    \item node has no children, ie.~is a virtual leaf (which corresponds to the size of the tree, \(n\));
    \item node has has children, ie.~is not a virtual leaf (which corresponds to \(n - 1\));
    \item node is a root (which corresponds to \(1\));
    \item node has its parent on the left, ie.~is right child of its parent;
    \item node has its parent on the right, ie.~is left child of its parent.
\end{itemize}

@@

With the proper interpretation, it may be possible to create sampling algorithm similar to Remy's one for binary trees\cite{remy,note}. It would have to take under consideration that some trees are being removed by other classes, so the probabilities would have to be set carrefuly.

@@

    \bibitem{remy}
    Laurent Alonso Alonso, Jean-Luc Remy, and René Schott,
    \textit{A Linear-Time Algorithm for the Generation of Trees},
    Algorithmica 17,
    1997,
    p. 162-182

>>>>

\(u, v\) are implicitly assumed to be nodes.

@@

We say that node \(v\) is an oneling \(\textiff\) \(v\) has no siblings. Note that the root is oneling even if it has no parent.

@@

@@

\(\n\) is implicitly defined as a default identifier of a node.

\(\no\) is implicitly defined as an identifier of non-existing node (eg.\ if \(v\) is a leaf, \(\tleft(v) = \n\)).

@@

We say that node \(v\) is an virtual-oneling \(\textiff\) \(v\) has no siblings being virtual nodes.

For binary tree, we say that node \(v\) is right (left) \(\textiff\) \(v\) is a right (left) child of its parent.

For binary tree, we say that node \(v\) has a right (left) parent \(u\) \(\textiff\) \(v\) is a left (right) child of \(u\).

@@

\(x_i\) is implicitly assumed to be an \(i\)-th element of the sequence \(X\), where \(x\) can be any lower-case letter and \(X\) is upper-case \(x\).

@@

We will be analyzing structures that can be expressed using recursive formulas. Eg.\ we can define full binary tree recursively as a node having no children (ie.\ being a leaf) or having exactly two subtrees being full binary trees. To express the alternative in the construction rule, we use \(+\). The reason will become obvious soon.

@@

There exist multiple well-known methods and tools that allow us to work with generating functions without the need of digging into details. The tool we will be using extensively in this work is Maple package called GFUN\cite{gfun}.

@@

\begin{definition}
    \label{def:holo_2}
    \cite{complexity}
    Equivalently to definition\ \ref{def:holo_1}, \(A = (a_i)_{i=0}^{\infty}\) is holonomic \(\textiff\) there exists generating function in the following form:
    \[\sum_{i=0}^{r} p_i(z) \gf{A}^{(i)}(z) = b(z)\]
    for some fixed \(r \in \N\); \(b, p_0, \ldots, p_r \in \poly\) not all identical to zero.
\end{definition}

@@

In the article we try to gather methods and results that may be helpful for anyone trying to find such interpretation, as we believe that such interpretation may give better insight into the nature of the structure and make it easier to study its behaviour.

@@

We define the size of a tree as a number of internal nodes (ie.\ nodes that are not leaves). Our generating function is therefore:

@@

This recurrence also describes Catalan numbers.

It is easy to see that tree of size \(n\) (ie. with \(n\) internal nodes) has \(n + 1\) leaves and therefore \(2n + 1\) nodes in total. Analogically the tree of size \(n + 1\) has \(n + 2\) leaves and \(2n + 3\) nodes in total. If you see some resemblance to our holonomic recurrence, you are perfectly right.

@@

First case can also be split into 2 subcases. Without loss of the generality we can analyze just one of the operations, let it be \(R_l\). The second case is symmetric.
\begin{itemize}
    \item for some nodes \(u \neq v\), \(R_l(T, u) = R_l(T, v)\). Let \(i_u, i_v\) be the indices of respectively \(\enc{u}\) and \(\enc{v}\) in \(\enc{T}\). Let \(\pointed{i}_u, \pointed{i}_v\) be the indices of pointed node respectively in \(\enc{R_l(T, u)}\) and \(\enc{R_l(T, v)}\). Due to the definition of \(R_l\), we get that \(\pointed{i}_u = i_u + 1\) and \(\pointed{i}_v = i_v + 1\), which implies that \(u = v\) what we assumed false.
    \item for some trees \(T \neq S\), \(R_l(T, u) = R_l(S, v)\). Let \(i_u, i_v\) be the indices of respectively \(\enc{u}\) in \(\enc{T}\) and \(\enc{v}\) in \(\enc{S}\). Let \(\pointed{i}_u, \pointed{i}_v\) be the indices of pointed node respectively in \(\enc{R_l(T, u)}\) and \(\enc{R_l(S, v)}\). If \(\pointed{i}_u \neq \pointed{i}_v\), this case is trivially false, so we can assume that \(\pointed{i}_u = \pointed{i}_v\). Therefore, again from the definition of \(R_l\), we have that \(i_u = i_v\). That implies that \(\enc{\tPre(T, u)} \enc{\tSub(u)} \enc{\tSuf(T, u)} = \enc{\tPre(S, v)} \enc{\tSub(v)} \enc{\tSuf(S, v)} \), which then sums up to \(\enc{T} = \enc{S}\), which we assumed false.
\end{itemize}

@@

The next case we will show is the case of Schröder trees\cite{schroder}.

@@

We can also move \(T(n)\) to the other side, ie.:

@@

Let's define 3 transformations, \(L_1\), \(L_2\) and \(R_1\):

\begin{center}
    \includeinlinesvg{.16}{schroder__lr_base}%
    \(\xRightarrow{L_1}\)%
    \includeinlinesvg{.16}{schroder__l1}%
    \hspace{.1\textwidth}%
    \includeinlinesvg{.16}{schroder__lr_base}%
    \(\xRightarrow{L_2}\)%
    \includeinlinesvg{.16}{schroder__l2}%

    \includeinlinesvg{.16}{schroder__lr_base}%
    \(\xRightarrow{R_1}\)%
    \includeinlinesvg{.16}{schroder__r1}%
\end{center}

@@

As you could see in the examples above, most of the solutions are based on the simple analysis of number of laves, internal nodes and other similar properties. In this section, we've gathered some methods that may be helpful when trying to analyze the problem of interpreting combinatorially some holonomic recurrence.

@@

First thing that we have to select is defining the size carefully. We have few standard choices of computing tree sizes:
\begin{itemize}
    \item counting all existing nodes,
    \item counting internal nodes,
    \item counting leaves,
    \item counting null-nodes (ie.\ children of the leaves),
    \item counting edges.
\end{itemize}

It may seem counter-intuitive at first, that this changes anything crucial, but with a change of the size definition, the holonomic recurrence may complicate itself drastically making the problem more difficult than it should be.

There are 2 most common symptoms that we may want to choose different method of counting the size:
\begin{itemize}
    \item we get empty classes of the objects, eg.\ if we analyze problem of full binary trees and define size as number of all existing nodes, we can not generate tree of even size, therefore half of our classes are empty;
    \item we want to join trees using common root and we get many \(\pm1\) in our notes. Note that if you count internal nodes, such catenation creates tree of size being sum of sizes of the components increased by 1. If you had defined the size as number of leaves or null-nodes, the size would be just plain sum of sizes of the smaller trees.
\end{itemize}

@@

Another method that may help finding the interpretation is inductive translation\cite{doron}. The method is based on the idea that most of the time there is an easy way to find some natural induction (not necessarily holonomic) that has trivial interpretation in bijective language.

@@

Using our inductive hypothesis, ie.\ holonomic recurrence and moving problematic cases out from the sums (in that case \(B(1)\) needs to be analyzed separately, as we did not define binary trees with no leaves), we can get the right side to use \(B(n - 1)\) and prove that the holonomic recurrence using non-linear one. The only part that is left is to translate this proof to the bijective language, which in such simple case as binary trees is almost trivial, especially if one has seen the solution once.

@@

If all elaborate methods has failed and the problem is still too big to interpret it by hand, a computer comes in handy.

@@

The definition above is quite popular, but introduces several problems. One of them is a problem of potential collision of lambda-abstracted variables and difficulty of finding to which lambda abstraction is the variable bound (or is it even free variable, ie.\ not bound). To address that, we can use de Bruijn indices.

@@

\begin{definition}
    \(s\) is a node with only right child.
\end{definition}

\begin{definition}
    \(l\) is a node with only left child.
\end{definition}

@@

We also created a framework\cite{transformator}, called transformator, that may help visualize our work. It is based on the idea that every transformation described using bijective language can be translated to a function transforming node and finite number of its surrounding nodes. This function can not discriminate nodes on the grounds of their kind. The only discrimination criteria that may occur may be connected to the following cases:
\begin{itemize}
    \item node has no children, ie.~is a virtual leaf (which corresponds to the size of the tree, \(n\));
    \item node has has children, ie.~is not a virtual leaf (which corresponds to \(n - 1\));
    \item node is a root (which corresponds to \(1\));
    \item node has its parent on the left, ie.~is right child of its parent;
    \item node has its parent on the right, ie.~is left child of its parent.
\end{itemize}

@@

With the proper interpretation, it may be possible to create sampling algorithm similar to Rémy's one for binary trees\cite{remy,note}. It would have to take under consideration that some trees are being removed by other classes, so the probabilities would have to be set carefully.

@@

    \bibitem{remy}
    Laurent Alonso Alonso, Jean-Luc Rémy, and René Schott,
    \textit{A Linear-Time Algorithm for the Generation of Trees},
    Algorithmica 17,
    1997,
    p. 162-182
